% main.tex -- Quantum Mechanics from Self-Modeling
% ASSERT_CONVENTION: sequential_product=a.b (dot notation), jordan_product=a circ b,
%   compression=C_p, axiom_source=arXiv:1803.11139 Definition 2 EXCLUSIVELY

\documentclass[11pt]{article}
\usepackage{preamble}

\title{Faithful Self-Modeling Is Complex Quantum Mechanics}

\author{Bryan Ehrlich\\[4pt]
  \normalsize Independent researcher, Trumbull, Connecticut 06611, USA\\
  \normalsize ehrlich.bryan@gmail.com}
\date{}

\begin{document}

\maketitle

\begin{abstract}
Faithful self-modeling \emph{is} complex quantum mechanics.  We
formalize self-modeling as four conditions on a finite-capacity
system (Definition~\ref{def:self-modeling-system})---none of which
mention complex numbers, Hilbert spaces, or Born probabilities---and
prove that every self-modeling system is isomorphic to $\Mnsa{n}$
with the L\"uders product.  The complex field, $C^{*}$-involution,
and full algebraic structure of quantum theory are consequences of
the definition, not inputs.  The proof builds a sequential product
from the system's own internal structure, verifies van de
Wetering's axioms to obtain Jordan algebra structure, then shows
that only complex systems admit the body--model composite the
definition requires---excluding every non-complex alternative.
Where other reconstruction programs require 3--6 independent axioms
and typically assume the complex field or local tomography as input,
here the complex field is derived and local tomography follows from
the internal-accessibility requirement built into the definition.  The downstream derivation chain (EJA classification through
$C^{*}$-promotion) is machine-verified in Lean~4 with Mathlib
(0~\texttt{sorry}); all 16 axioms cite published theorems of
Alfsen--Shultz and van de Wetering, and the novel construction of
Sections~3--4 is fully proved, not axiomatized.
\end{abstract}


%% =====================================================================
%%  SECTION 1: INTRODUCTION
%% =====================================================================
\section{Introduction}
\label{sec:intro}

Faithful self-modeling and complex quantum mechanics are the same
thing.  We formalize self-modeling as four conditions on a
finite-dimensional system---spectral state space, faithful internal
tracking, minimal body--model composite, and irreducibility---none
of which mention complex numbers, Hilbert spaces, or $C^{*}$-algebras.
We then prove that every system satisfying these conditions is
isomorphic to $\Mnsa{n}$, the self-adjoint $n \times n$ complex
matrices equipped with the L\"uders sequential product
$\seqp{a}{b} = \sqrt{a}\, b\, \sqrt{a}$ and the conjugate-transpose
involution $X^{*} = X^{\dagger}$.  The complex field, Born-rule
structure, and $C^{*}$-involution all emerge as consequences of the
self-modeling definition.

This result contributes to the quantum reconstruction program, which
seeks to derive the mathematical framework of quantum theory from
physically transparent principles.  Beginning with Hardy's five-axiom
reconstruction~\cite{Hardy2001}, the program has produced a succession
of results that characterize quantum theory within the landscape of
generalized probabilistic theories (GPTs): Chiribella, D'Ariano, and
Perinotti~\cite{CDP2011} used six axioms including Purification;
Daki\'c and Brukner~\cite{DakicBrukner2011} used four axioms
including local tomography; Masanes and
M\"uller~\cite{MasanesMueller2011,MasanesMueller2013} distilled the
framework to three postulates assuming tomographic locality; and
Barnum, M\"uller, and
Ududec~\cite{BarnumMuellerUdudec2014} used four postulates including
energy observability.  A comprehensive review of the landscape is
given by M\"uller~\cite{Mueller2021}.  A common feature of these
programs is the use of multiple independent operational axioms, and in
most cases the complex number field---or a property that selects it,
such as local tomography---is assumed rather than derived.

Our main result can be stated informally as follows.

\begin{theorem}[Main result, informal]\label{thm:main-informal}
  Every self-modeling system
  (Definition~\ref{def:self-modeling-system}) is isomorphic to
  $\Mnsa{n}$ for some $n \geq 2$, equipped with the L\"uders
  sequential product and conjugate-transpose involution.
\end{theorem}

\noindent
The converse is immediate: $\Mnsa{n}$ admits a faithful self-model
(take $\varphi = \mathrm{id}$, with the standard tensor product as
composite).  The result is therefore a \emph{characterization}: among
finite-dimensional spectral order unit spaces, self-modeling systems
are exactly the complex matrix algebras.  The content of the theorem
is the forward direction---that four conditions about self-modeling,
none of which mention complex numbers or Hilbert spaces, turn out to
characterize exactly the systems that have them.

The derivation proceeds through a chain of implications:
self-modeling $\to$ sequential product construction $\to$ axiom
verification $\to$ Euclidean Jordan algebra classification $\to$
local tomography $\to$ type exclusion $\to$ $C^{*}$-algebra promotion
$\to$ involution exhibition.
\Cref{fig:chain} illustrates this chain, distinguishing the novel
contributions of this paper from the published theorems invoked.

\Cref{tab:comparison} compares the premise count and complex-selection
mechanism of this work against the major reconstruction programs.

\begin{table}[ht]
\centering
\caption{Comparison of quantum reconstruction programs.  $\mathbb{C}$:
complex field; Born: Born-rule structure; $C^{*}$: full
$C^{*}$-algebraic structure including involution.  A = assumed,
D = derived.  Framework: GPT = generalized probabilistic theory,
OPT = operational-probabilistic theory, OUS = order unit space with
Alfsen--Shultz compressions.  Our
Definition~\ref{def:self-modeling-system} packages four conditions
(see text); the richer OUS framework is part of the input.}
\label{tab:comparison}
\smallskip\small
\begin{tabular}{@{}llcccc@{}}
\toprule
Program & Input & $\mathbb{C}$ & Born & $C^{*}$ & Framework \\
\midrule
Hardy~\cite{Hardy2001} & 5 axioms & A & D & A & GPT \\
CDP~\cite{CDP2011} & 6 axioms & D & D & A & OPT \\
Masanes--M\"uller~\cite{MasanesMueller2013} & 3 postulates & A & D & A & GPT \\
BMU~\cite{BarnumMuellerUdudec2014} & 4 postulates & D & D & A & GPT \\
\textbf{This work} & \textbf{Def.~\ref{def:self-modeling-system} (4 conds.)} & \textbf{D} & \textbf{D} & \textbf{D} & \textbf{OUS} \\
\bottomrule
\end{tabular}
\end{table}

% Derivation chain figure
\input{figures/derivation-chain}

The paper is organized as follows.
\Cref{sec:prelim} introduces order unit spaces, compressions, Peirce
decompositions, sequential products, and the self-modeling definition.
\Cref{sec:sp} presents the main novel construction: the self-modeling
sequential product, the naive extension and its failure, the corrected
product via Peirce feedback, and the determination of the mixing
function.
\Cref{sec:axioms} verifies that the corrected product satisfies
axioms S1--S7 of van de Wetering~\cite{vandeWetering2019b}.
\Cref{sec:composite} constructs the body--model composite, proves
local tomography, and addresses the entangled sector.
\Cref{sec:exclusion} excludes all non-complex Euclidean Jordan algebra
types and promotes the algebra to a $C^{*}$-algebra via the theorems
of van de Wetering, Barnum--Wilce, and Hanche-Olsen.
\Cref{sec:discussion} discusses the scope and limitations of the
result.

We close this introduction with an honest statement of scope.
Definition~\ref{def:self-modeling-system} formalizes ``self-modeling
system'' as four conditions, each unpacking one word of the phrase:
\emph{system} (finite-dimensional, spectral),
\emph{modeling} (faithful tracking),
\emph{self-}\ (internal, minimal composite), and
\emph{the whole} (simple, irreducible).
Remove any condition and the concept of self-modeling breaks; add
any condition and you assume more than self-modeling requires.
\Cref{subsec:unified} develops this argument in detail.  The
definition characterizes \emph{composable} self-modelers (subsystems
that can participate in composites); the non-composable case (the
``self-modeling basin'') is treated in a companion paper~\cite{ehrlich2026sm}.


%% =====================================================================
%%  SECTION 2: PRELIMINARIES
%% =====================================================================
\section{Preliminaries}
\label{sec:prelim}


\subsection{Order Unit Spaces}
\label{sec:ous}

We work within the framework of order unit spaces, following
Alfsen--Shultz~\cite{AlfsenShultz2003}.

\begin{definition}[Order unit space]\label{def:ous}
  An \emph{order unit space} $(V, V^{+}, \id)$ is a real ordered
  vector space with a proper cone $V^{+}$ (closed, convex, pointed)
  generating $V$, equipped with a distinguished Archimedean order unit
  $\id$: for every $a \in V$ there exists $r \geq 0$ with
  $-r\id \leq a \leq r\id$.  The \emph{order unit norm} is
  $\|a\| = \inf\{r \geq 0 : -r\id \leq a \leq r\id\}$.
\end{definition}

The \emph{effect space} of $V$ is
$\eff{V} = \{a \in V : 0 \leq a \leq \id\}$.  An effect
$p \in \eff{V}$ is \emph{sharp} (a \emph{projective unit}) if $p$ and
$\id - p$ are both extremal in $\eff{V}$, or equivalently if $p$
admits no non-trivial refinement.  We write $\Proj(V)$ for the set of
projective units.

Every effect in a spectral order unit space admits a \emph{spectral
decomposition} $a = \sum_{i} \lambda_{i}\, p_{i}$, where the $p_{i}$
are mutually orthogonal projective units with
$\sum_{i} p_{i} = \id$ and the $\lambda_{i}$ are distinct elements of
$[0,1]$.  This decomposition exists as a property of the OUS lattice
structure~\cite{AlfsenShultz2003}, prior to and independent of any
sequential product.

% Clause (i) of Definition~\ref{def:self-modeling-system}.
% Label retained for backward compatibility with cross-references.
\phantomsection\label{ass:finite}%


\subsection{Compressions and Peirce Decomposition}
\label{sec:compressions}

For each projective unit $p \in \Proj(V)$, the Alfsen--Shultz theory
provides a \emph{compression} $\comp{p} : V \to V$, a positive linear
projection satisfying $\comp{p}(\id) = p$.  The compression $\comp{p}$
projects onto the face generated by $p$, and it plays the role of the
L\"uders operation $b \mapsto p\, b\, p$ in the general OUS setting.

\begin{definition}[Peirce decomposition]\label{def:peirce}
  For a projective unit $p \in \Proj(V)$, the \emph{Peirce
  decomposition} of $V$ relative to $p$ is the direct-sum
  decomposition
  \begin{equation}\label{eq:peirce}
    V = V_{2}(p) \oplus V_{1}(p) \oplus V_{0}(p),
  \end{equation}
  where $V_{2}(p) = \mathrm{range}(\comp{p})$ is the face of $p$,
  $V_{0}(p) = \mathrm{range}(\comp{p^{\perp}})$ is the face of
  $p^{\perp} = \id - p$, and
  $V_{1}(p) = \ker(\comp{p}) \cap \ker(\comp{p^{\perp}})$ is the
  \emph{Peirce $1$-space}.
\end{definition}

A key identity is:
\begin{equation}\label{eq:pinching}
  \comp{p} + \comp{p^{\perp}} = \pinch_{p}
  \quad (\text{the pinching map, \emph{not} the identity}).
\end{equation}
The pinching map annihilates $V_{1}(p)$.  In the commutative
(classical) case $V_{1}(p) = \{0\}$, so the pinching map is the
identity; in the non-commutative case $V_{1}(p) \neq \{0\}$ and the
identity fails.  This distinction drives the entire construction in
\Cref{sec:sp}.

More generally, for a resolution of unity
$\{p_{1}, \ldots, p_{n}\}$ (mutually orthogonal projective units with
$\sum_{i} p_{i} = \id$), the \emph{Peirce $(i,j)$ projection}
$\peirce{ij}$ extracts the component of $b$ in the $(i,j)$ Peirce
$1$-space $V_{1}(p_{i}, p_{j})$:
\begin{equation}\label{eq:peirce-proj}
  \peirce{ij}(b) \in V_{1}(p_{i}, p_{j}),
  \qquad
  b = \sum_{i} \comp{p_{i}}(b) + \sum_{i < j} \peirce{ij}(b).
\end{equation}


\subsection{Sequential Products}
\label{sec:sp-axioms}

Following Gudder and Greechie~\cite{GudderGreechie2002} and van de
Wetering~\cite{vandeWetering2019b}, we define a sequential product on
an order unit space.

\begin{definition}[Sequential product space,
  {\cite[Def.~2]{vandeWetering2019b}}]
\label{def:sp}
  Let $(V, \leq, \id)$ be an order unit space equipped with a binary
  operation $(\cdot) : \eff{V} \times \eff{V} \to \eff{V}$.  Write
  $a \compatible b$ (``$a$ and $b$ are compatible'') when
  $\seqp{a}{b} = \seqp{b}{a}$.  Then $(V, \cdot)$ is a \emph{sequential
  product space} if the following axioms hold:
  \begin{enumerate}[label=\textup{(S\arabic*)},leftmargin=3em]
    \item\label{ax:S1} \textbf{Additivity:}
      $\seqp{a}{(b+c)} = \seqp{a}{b} + \seqp{a}{c}$ whenever
      $b + c \leq \id$.
    \item\label{ax:S2} \textbf{Continuity:}
      The map $a \mapsto \seqp{a}{b}$ is continuous in the order unit
      norm.
    \item\label{ax:S3} \textbf{Unitality:}
      $\seqp{\id}{a} = a$ for all $a \in \eff{V}$.
    \item\label{ax:S4} \textbf{Symmetry of orthogonality:}
      If $\seqp{a}{b} = 0$ then $\seqp{b}{a} = 0$.
    \item\label{ax:S5} \textbf{Associativity of compatible effects:}
      If $a \compatible b$ then
      $\seqp{a}{(\seqp{b}{c})} = \seqp{(\seqp{a}{b})}{c}$.
    \item\label{ax:S6} \textbf{Additivity of compatibility:}
      If $a \compatible b$ then $a \compatible (\id - b)$; if also
      $a \compatible c$ then $a \compatible (b + c)$ whenever
      $b + c \leq \id$.
    \item\label{ax:S7} \textbf{Multiplicativity of compatibility:}
      If $a \compatible b$ and $a \compatible c$ then
      $a \compatible \seqp{b}{c}$.
  \end{enumerate}
\end{definition}

Two central results of van de Wetering are used as black boxes
throughout this paper.

\begin{theorem}[{\cite[Thm.~1]{vandeWetering2019b}}]\label{thm:vdw1}
  A finite-dimensional sequential product space is order-isomorphic
  to a Euclidean Jordan algebra.
\end{theorem}

\begin{theorem}[{\cite[Thm.~3]{vandeWetering2019b}}]\label{thm:vdw3}
  If a sequential product space has a locally tomographic composite
  that also carries a sequential product, then the sequential product
  space embeds into a $C^{*}$-algebra.
\end{theorem}

The Jordan--von Neumann--Wigner
classification~\cite{JordanVonNeumannWigner1934} of simple
finite-dimensional Euclidean Jordan algebras yields five types:
$\Mnsa{n}$ over $\mathbb{R}$, $\mathbb{C}$, or $\mathbb{H}$; the
spin factors $\spinfactor{n}$; and the exceptional Albert algebra
$M_{3}(\mathbb{O})\sa$.  Our task is to show that self-modeling
is complex quantum mechanics.


\subsection{Self-Modeling Systems}
\label{sec:self-modeling}

We now introduce the central definition.

Consider a physical system $B$ (the ``body'') described by a
finite-dimensional spectral OUS $(V_{B}, V_{B}^{+}, \id_{B})$.  A
\emph{self-model} of $B$ is a second system $M$ (the ``model''),
described by an OUS $(V_{M}, V_{M}^{+}, \id_{M})$, together with a
positive unital map $\varphi: V_{B} \to V_{M}$ (the \emph{tracking
map}) that encodes the operational cycle:
\begin{enumerate}[label=(\roman*)]
  \item \emph{Prepare:} the body $B$ is in some state;
  \item \emph{Update:} testing a sharp effect $p$ on $B$ triggers
    $M$ to undergo the compression $\comp{\varphi(p)}$, updating
    the model;
  \item \emph{Measure:} a second effect $b$ is tested on $B$.
\end{enumerate}
The composite outcome of this cycle---``first $p$, then $b$''---defines
a binary operation on the effect space of $B$.

\begin{definition}[Self-modeling system]\label{def:self-modeling-system}
  A \emph{self-modeling system} is a tuple $(V, \varphi, V_{BM})$
  satisfying:
  \begin{enumerate}[label=(\roman*)]
    \item\label{sms:finite} $V$ is a nontrivial finite-dimensional
      spectral order unit space (i.e., $V$ has at least two
      orthogonal nontrivial projective units);
    \item\label{sms:faithful} $\varphi: V_{B} \to V_{M}$ is an
      order isomorphism (faithful tracking);
    \item\label{sms:minimal} $V_{BM}$ is the minimal composite OUS
      carrying product states, product effects, non-signaling
      constraints, and product-form sequential product;
    \item\label{sms:simple} $V$ has no nontrivial direct-sum
      decomposition: there is no splitting $V = V_1 \oplus V_2$ into
      nonzero order-unit subspaces.
  \end{enumerate}
\end{definition}

Each clause unpacks one word of the concept:
\begin{itemize}[nosep]
  \item \emph{System} \ref{sms:finite}: has states, measurements,
    distinguishable outcomes.  Finite because a self-modeler in a
    finite region has finite information
    capacity~\cite{Bekenstein1981}.  Nontriviality excludes the
    degenerate case $V \cong \mathbb{R}$ (no nontrivial projective
    units, vacuous Peirce structure): a system with nothing to
    distinguish has nothing to model.  All non-complex types with
    nontrivial projective units are excluded by the local
    tomography argument (Theorem~\ref{thm:local-tomo}), not by the
    definition.
  \item \emph{Modeling} \ref{sms:faithful}: the internal model is
    accurate---an order isomorphism preserves all operationally
    accessible structure.
  \item \emph{Self-}\ \ref{sms:minimal}: the model is internal.
    The only measurements available to the self-modeler are product
    measurements on the body and model.  If the body--model composite
    carried states that product measurements could not distinguish,
    those states would be opaque to the self-modeler---aspects of its
    own composite that it cannot probe or model.  Minimality says:
    the composite has no such hidden structure.  The product-form
    sequential product (Definition~\ref{def:product-sp}) encodes
    non-signaling: testing the body and updating the model are
    independent operations on separate subsystems.
  \item \emph{The whole} \ref{sms:simple}: a direct-sum algebra
    decomposes into sectors invisible to each other; a system in one
    sector cannot model another.  Simplicity is the algebraic
    expression of ``whole models whole.''
\end{itemize}
Remove any clause and the concept of self-modeling breaks.  Add any
clause and you have assumed more than self-modeling requires.

\begin{remark}
  Approximate or partial self-models (where $\varphi$ is merely
  positive or injective but not surjective) might yield weaker
  algebraic structures; we do not pursue this here.
  Definition~\ref{def:self-modeling-system} characterizes
  \emph{composable} self-modelers---subsystems that can participate
  in composites.  The non-composable case
  ($h_{3}(\mathbb{O})$, which admits no composite at all) is treated
  in a companion paper~\cite{ehrlich2026sm}.
\end{remark}


%% =====================================================================
%%  SECTION 3: THE SELF-MODELING SEQUENTIAL PRODUCT
%% =====================================================================
\section{The Self-Modeling Sequential Product}
\label{sec:sp}


\subsection{From Self-Modeling to Sequential Product}
\label{sec:sp-sharp}

The self-modeling cycle of \Cref{sec:self-modeling} produces a natural
binary operation on the effects of $B$.  For sharp effects
$p, q \in \Proj(V_{B})$, the composite outcome ``first test $p$,
update the model, then test $q$'' is encoded by the compression
product:
\begin{equation}\label{eq:sharp-product}
  \seqp{p}{q} = \comp{p}(q),
\end{equation}
where $\comp{p}$ is the Alfsen--Shultz compression for the face
generated by $p$ in $V_{B}$.

Several properties of this product follow immediately from the
compression axioms.  It maps effects to effects (positivity of
$\comp{p}$).  It satisfies $\seqp{\id}{q} = q$ (since $\comp{\id}$ is
the identity).  On a simplex (classical state space), it reduces to
the pointwise product
$\seqp{p}{q} = (p_{1}q_{1}, \ldots, p_{n}q_{n})$, matching the unique
sequential product on simplices~\cite{GudderGreechie2002}.

\begin{remark}
  The product~\eqref{eq:sharp-product} is defined entirely on
  $V_{B}$ using $B$'s own compression structure.  The tracking map
  $\varphi$ provides the physical justification---testing $p$ on $B$
  causes the model $M$ to undergo $\comp{\varphi(p)}$, which feeds
  back to the context for testing $q$---but the algebraic product is
  intrinsic to $B$.
\end{remark}


\subsection{The Naive Extension and Its Failure}
\label{sec:sp-naive}

To extend the product from sharp to general effects, one attempts a
spectral extension.  Given $a = \sum_{i} \lambda_{i}\, p_{i}$
(spectral decomposition in the OUS), define:
\begin{equation}\label{eq:naive}
  \seqp{a}{b} \stackrel{?}{=}
    \sum_{i} \lambda_{i}\, \comp{p_{i}}(b).
\end{equation}
This is the natural extension: the effect $a$ acts on $b$ as a
weighted sum of compressions, one for each spectral projector.

However, this extension \emph{fails} axiom \ref{ax:S3} (unitality)
on non-commutative order unit spaces.  To see why, set $a = \id$ with
spectral decomposition $\id = \sum_{i} p_{i}$ (all eigenvalues equal
to~$1$).  Then~\eqref{eq:naive} gives
\begin{equation}\label{eq:naive-failure}
  \seqp{\id}{b}
  = \sum_{i} \comp{p_{i}}(b)
  = \pinch_{p_{1}, \ldots, p_{n}}(b)
  \neq b
  \quad\text{in general.}
\end{equation}
The sum of compressions is the \emph{pinching map}, not the identity.
By the Peirce decomposition~\eqref{eq:peirce-proj}, the pinching map
annihilates all Peirce $1$-space components: $\peirce{ij}(b)$ is
discarded for every $i < j$.  In the commutative case
$V_{1} = \{0\}$, the pinching map \emph{is} the identity and the
extension succeeds---this is why the classical limit works.  In the
non-commutative case, the Peirce $1$-space is nontrivial, and the
naive extension destroys off-diagonal information.

This failure is not a bug; it reveals a structural gap.  The
compression sum captures the Peirce $2$-space (diagonal) contributions
correctly but provides no prescription for the Peirce $1$-space
(off-diagonal) terms.  Closing this gap requires new input.


\subsection{The Corrected Product via Peirce Feedback}
\label{sec:sp-corrected}

The missing Peirce $1$-space contribution comes from the self-modeling
feedback loop.  We now show that the form of the corrected product is
\emph{forced} by the axioms, not chosen.  Any binary operation on
effects satisfying \ref{ax:S1} (additivity in the second argument),
\ref{ax:S2} (continuity in the first), and the sharp constraint
$\seqp{p}{b} = \comp{p}(b)$ must respect the Peirce decomposition:
the map $\seqp{a}{\cdot}$ is a linear endomorphism (by \ref{ax:S1}
and finite-dimensionality) whose range on each Peirce subspace is
constrained to that subspace (by the compression algebra structure).
The diagonal coefficients are pinned to the eigenvalues by
eigenvalue coalescence and \ref{ax:S2}.  Thus the general product
for $a = \sum_{i} \lambda_{i}\, p_{i}$ must take the form:
\begin{equation}\label{eq:general-product}
  \seqp{a}{b}
  = \sum_{i} \lambda_{i}\, \comp{p_{i}}(b)
  + \sum_{i < j} f(\lambda_{i}, \lambda_{j})\, \peirce{ij}(b),
\end{equation}
where the first sum captures diagonal (Peirce $2$-space) contributions
via compressions, and the second sum restores the off-diagonal (Peirce
$1$-space) contributions through a \emph{mixing function}
$f: [0,1] \times [0,1] \to \mathbb{R}$ to be determined.

\emph{Why this form is forced.}
By \ref{ax:S1}, $\seqp{a}{\cdot}$ is a linear endomorphism of~$V$.
The Peirce decomposition with respect to the spectral projectors of~$a$
decomposes $V$ into subspaces $V_2(p_i)$ and $V_1(p_i, p_j)$.
Compressions project onto these subspaces
(Alfsen--Shultz~\cite{AlfsenShultz2003}), so linearity gives a block
decomposition: $\seqp{a}{\cdot}$ maps each Peirce subspace to itself.
On $V_2(p_i)$, the sharp constraint $\seqp{p_i}{b} = \comp{p_i}(b)$
and coalescence ($\lambda_i \to \lambda_j$ via \ref{ax:S2}) fix the
action to $\lambda_i \comp{p_i}(\cdot)$.  On $V_1(p_i, p_j)$, the
action is a linear endomorphism $E_{ij}(\lambda_i, \lambda_j)$ of
$V_1(p_i, p_j)$, depending continuously on the eigenvalues by
\ref{ax:S2}.  In Proposition~\ref{prop:coherence}, we prove that
compatible associativity (\ref{ax:S5}), eigenvalue symmetry, and
coalescence force $E_{ij}$ to be a scalar multiple of the identity
on each $V_1(p_i, p_j)$: that is,
$E_{ij}(\lambda_i, \lambda_j) = f(\lambda_i, \lambda_j) \cdot
\peirce{ij}$ for a single mixing function~$f$, simultaneously
determining $f = \sqrt{\lambda_i\, \lambda_j}$.
Thus~\eqref{eq:general-product} is the unique form compatible with
\ref{ax:S1}--\ref{ax:S5} and the sharp constraint.

The mixing function is constrained by the requirement that the product
map effects to effects: $0 \leq \seqp{a}{b} \leq \id$ for all effects
$a, b$.  This yields a \emph{positivity bound}.

\begin{proposition}[Positivity bound]\label{prop:pos-bound}
  For the product~\eqref{eq:general-product} to map
  $\eff{V} \times \eff{V}$ into $\eff{V}$, the mixing function must
  satisfy
  \begin{equation}\label{eq:pos-bound}
    |f(\lambda_{i}, \lambda_{j})| \leq
    \sqrt{\lambda_{i}\, \lambda_{j}}
  \end{equation}
  for all $\lambda_{i}, \lambda_{j} \in [0,1]$.
\end{proposition}

\begin{proof}
  Every pair of orthogonal atoms $p_{i}, p_{j}$ in a spectral OUS
  generates a two-level face isomorphic to a spin factor, on which
  the Schur complement criterion gives
  $|f(\lambda_{i}, \lambda_{j})| \leq \sqrt{\lambda_{i}\,
  \lambda_{j}}$.  Face effects embed into global effects, so the
  bound lifts to the full space.
\end{proof}

For the saturating choice $f = \sqrt{\lambda_i\, \lambda_j}$
(Proposition~\ref{prop:coherence}), the product maps effects to
effects globally: the coefficient matrix
$[M_{ij}]$ with $M_{ij} = \sqrt{\lambda_i\, \lambda_j}$ factors as
$vv^{T}$ with $v_i = \sqrt{\lambda_i}$, hence is positive
semidefinite.  Together with positivity of compressions, this gives
$\seqp{a}{b} \geq 0$ when $a, b \geq 0$; the upper bound
$\seqp{a}{b} \leq \id$ follows from \ref{ax:S1} (additivity in~$b$)
and \ref{ax:S3} (unitality), both verified in \Cref{sec:axioms}.


\subsection{Determining the Mixing Function}
\label{sec:sp-faithful}

As shown in Section~\ref{sec:sp-corrected}, the self-modeling
construction provides a product whose action on each Peirce
$1$-space $V_1(p_i, p_j)$ is an undetermined linear endomorphism
$E_{ij}(\lambda_i, \lambda_j)$.  We now show that internal
consistency forces $E_{ij}$ to be scalar and uniquely
determines the scalar.

\begin{proposition}[Scalar action and maximal coherence]\label{prop:coherence}
  The action of $\seqp{a}{\cdot}$ on each Peirce $1$-space
  $V_1(p_i, p_j)$ is scalar multiplication by
  $\sqrt{\lambda_i\, \lambda_j}$:
  \begin{equation}\label{eq:faithful-f}
    E_{ij}(\lambda_i, \lambda_j)
    = \sqrt{\lambda_{i}\, \lambda_{j}} \cdot \mathrm{Id}_{V_1(p_i, p_j)}.
  \end{equation}
  In particular, the mixing function is
  $f(\lambda_i, \lambda_j) = \sqrt{\lambda_i\, \lambda_j}$
  and the product has the form~\eqref{eq:general-product}.
\end{proposition}

\begin{proof}
  The argument treats \ref{ax:S5} as a \emph{consistency constraint}:
  we seek the endomorphisms $E_{ij}$ for which the resulting product
  satisfies \ref{ax:S5}.
  (Once $E_{ij}$ is determined, \Cref{sec:axioms} verifies S1--S7 in
  full; the logic is not circular.)

  \emph{Step 1: Multiplicative composition.}
  For compatible $a = \alpha_1 p_1 + \alpha_2 p_2$ and
  $b = \beta_1 p_1 + \beta_2 p_2$ sharing an eigenbasis, imposing
  \ref{ax:S5}---i.e., requiring
  $\seqp{a}{(\seqp{b}{d})} = \seqp{(\seqp{a}{b})}{d}$ for all
  effects~$d$---and comparing the action on $V_1(p_1, p_2)$ gives
  \begin{equation}\label{eq:endo-compose}
    E(\alpha_1, \alpha_2) \circ E(\beta_1, \beta_2)
    = E(\alpha_1 \beta_1,\; \alpha_2 \beta_2),
  \end{equation}
  where $E = E_{12}$ and $\circ$ denotes composition of
  endomorphisms of $V_1(p_1, p_2)$.

  \emph{Step 2: Factorization.}
  Define $h(x) = E(x, 1)$ and $k(y) = E(1, y)$.
  Equation~\eqref{eq:endo-compose} gives $h(x_1)h(x_2) = h(x_1 x_2)$
  and $k(y_1)k(y_2) = k(y_1 y_2)$, so $h$ and $k$ are continuous
  multiplicative maps $(0,1] \to \mathrm{End}(V_1)$.  Setting
  $\alpha_2 = \beta_1 = 1$ gives
  $E(x, y) = h(x)\, k(y)$, and $h$, $k$ commute (both orderings
  give $E(x, y)$).  Every continuous multiplicative homomorphism
  $(0,1] \to \mathrm{GL}(W)$ on a finite-dimensional space~$W$ has the
  form $t \mapsto t^{A} = \exp(A \ln t)$ for some
  $A \in \mathrm{End}(W)$ (Acz\'el~\cite{Aczel1966}, Ch.~4).
  Applied to $h$ and~$k$:
  $h(x) = x^{A}$ and $k(y) = y^{B}$ for matrices $A$, $B$ with
  $[A, B] = 0$.

  \emph{Step 3: Symmetry.}
  The Peirce $1$-space $V_1(p_1, p_2) = V_1(p_2, p_1)$ is the same
  subspace regardless of index order.
  The product sums over unordered pairs, so
  $E(x, y) = E(y, x)$ as endomorphisms.  This gives
  $x^A\, y^B = y^A\, x^B$ for all $x, y > 0$, which forces $A = B$.
  Hence $E(x, y) = (xy)^A$.

  \emph{Step 4: Coalescence.}
  By \ref{ax:S2}, as $\lambda_1 \to \lambda_2 = \lambda$, the product
  reduces to $\seqp{(\lambda\id)}{b} = \lambda b$, giving
  \begin{equation}\label{eq:coalescence}
    E(\lambda, \lambda) = \lambda \cdot \mathrm{Id}
    \qquad \text{for all } \lambda \in (0,1].
  \end{equation}
  Substituting $E(\lambda, \lambda) = (\lambda^2)^A = \lambda^{2A}$
  and comparing with $\lambda\, \mathrm{Id} = \lambda^{\mathrm{Id}}$
  gives $2A = \mathrm{Id}$, so $A = \tfrac{1}{2}\, \mathrm{Id}$.
  Therefore
  $E(x, y) = (xy)^{\mathrm{Id}/2}
  = \sqrt{xy}\, \mathrm{Id}$.

  The endomorphism is scalar, with
  $f(\lambda_i, \lambda_j) = \sqrt{\lambda_i\, \lambda_j}$.
\end{proof}

\noindent
To clarify the logical structure: Section~\ref{sec:sp-corrected}
shows that linearity, continuity, and the sharp constraint force
the product into a block-diagonal form with an undetermined
endomorphism on each Peirce $1$-space.  The present proposition
shows that \emph{requiring} \ref{ax:S5} (a necessary condition for
any sequential product space) forces the endomorphism to be scalar
and uniquely determines the scalar.
Section~\ref{sec:axioms} then verifies all of S1--S7 for the
resulting product with
$f = \sqrt{\lambda_{i}\, \lambda_{j}}$, closing the loop.
The role of self-modeling is constructive: the tracking
isomorphism~$\varphi$ provides the product on which the axioms
act.  Without~$\varphi$, there is no product to constrain; once
the product exists, the axioms determine~$f$ uniquely.

The corrected product with
$f = \sqrt{\lambda_{i}\, \lambda_{j}}$ is:
\begin{equation}\label{eq:corrected-product}
  \boxed{
  \seqp{a}{b}
  = \sum_{i} \lambda_{i}\, \comp{p_{i}}(b)
  + \sum_{i < j} \sqrt{\lambda_{i}\, \lambda_{j}}\,
    \peirce{ij}(b).
  }
\end{equation}

\begin{remark}
  On $\Mnsa{2}$, the corrected
  product~\eqref{eq:corrected-product} coincides with the L\"uders
  product $\seqp{a}{b} = \sqrt{a}\, b\, \sqrt{a}$.  This equivalence
  is a \emph{consequence} of the construction, not a premise.  The
  L\"uders formula was never assumed; it emerges from the positivity
  bound and faithful self-modeling.
\end{remark}

\begin{remark}[Division of labor]\label{rem:division}
  The tracking map~$\varphi$ and the sequential-product axioms play
  complementary roles.  Self-modeling provides the operational
  justification for a binary product on effects (the
  test--update--test cycle) and constrains its form to the
  ansatz~\eqref{eq:general-product}.  The axioms---specifically
  \ref{ax:S5}---then pin the mixing function uniquely.  Neither
  ingredient suffices alone: without~$\varphi$, there is no product
  to constrain; without~\ref{ax:S5}, the mixing function is
  underdetermined.  Heuristically, $f = 0$ corresponds to a trivial
  (non-faithful) model and $f = \sqrt{\lambda_{i}\, \lambda_{j}}$
  to a faithful one, but we do not prove a quantitative theorem
  linking the fidelity of~$\varphi$ to the value of~$f$; the
  connection is that faithfulness of~$\varphi$ justifies seeking the
  product that saturates the positivity
  bound~\eqref{eq:pos-bound}, and \ref{ax:S5} identifies that
  product uniquely.
\end{remark}

The remaining clauses of
Definition~\ref{def:self-modeling-system}---minimality
\ref{sms:minimal} and simplicity \ref{sms:simple}---enter
in \Cref{sec:composite}.
% Labels retained for backward compatibility with cross-references.
\phantomsection\label{ass:faithful}%
\phantomsection\label{ass:minimal}%
\phantomsection\label{ass:simple}%


\subsection{Circularity Check}
\label{sec:circularity}

A natural concern is whether the scalar function
$\sqrt{\lambda_{i}\, \lambda_{j}}$
in~\eqref{eq:corrected-product} imports $C^{*}$-algebraic structure
through the back door.  It does not.  The square root in the mixing
function operates on \emph{real eigenvalues}
$\lambda_{i}, \lambda_{j} \in [0,1]$: it is arithmetic, specifically
$\sqrt{x} : [0,1] \to [0,1]$, the unique non-negative real number
whose square is $x$.  No operator square root, no trace, no inner
product, no density matrix, and no Hilbert space structure is used in
the construction.  The only mathematical objects employed are:
\begin{itemize}
  \item the order unit space $(V, V^{+}, \id)$;
  \item the effect space $\eff{V}$ and its projective units;
  \item Alfsen--Shultz compressions $\comp{p}$ for faces of
    projective units;
  \item the Peirce decomposition~\eqref{eq:peirce-proj} derived from
    compressions;
  \item spectral decomposition of effects in the OUS;
  \item the scalar function $\sqrt{\cdot}$ on
    $[0,1] \subset \mathbb{R}$.
\end{itemize}
Complex numbers, Hilbert spaces, and $C^{*}$-algebraic operations do
not appear until the conclusion (\Cref{sec:exclusion}), where they
emerge as consequences of the classification theorems.


%% =====================================================================
%%  SECTIONS 4--6: BODY (from Plan 02)
%% =====================================================================

\input{sections/axiom-verification}

\input{sections/composite-lt}

\input{sections/type-exclusion}


%% =====================================================================
%%  SECTION 7: DISCUSSION (from Plan 03)
%% =====================================================================

\input{sections/discussion}


%% =====================================================================
%%  ACKNOWLEDGMENTS
%% =====================================================================

\section*{Acknowledgments}

The author thanks John van de Wetering for making his sequential
product theorems publicly available and clearly stated, and the
anonymous referees in advance for their scrutiny.

\paragraph{AI disclosure.}
This work used AI tools throughout: Claude (Anthropic) for
exploratory mathematical conversations, literature search, manuscript
drafting, and assistance with the Lean~4 formalization; GPT-5.4
(OpenAI) for adversarial review of mathematical claims.  All
definitions, theorems, proofs, and scientific conclusions are the
author's.  The author takes full responsibility for the correctness
and originality of the mathematical content.


%% =====================================================================
%%  APPENDICES
%% =====================================================================
\appendix

\input{sections/appendix-proofs}

\input{sections/appendix-numerical}


%% =====================================================================
%%  BIBLIOGRAPHY
%% =====================================================================
\bibliographystyle{alpha}
\bibliography{refs}

\end{document}
