% main.tex -- Quantum Mechanics from Self-Modeling
% ASSERT_CONVENTION: sequential_product=a.b (dot notation), jordan_product=a circ b,
%   compression=C_p, axiom_source=arXiv:1803.11139 Definition 2 EXCLUSIVELY

\documentclass[11pt]{article}
\usepackage{preamble}

\title{Quantum Mechanics from Self-Modeling:\\
Deriving Complex $C^*$-Algebraic Structure\\from a Single Operational Premise}

\author{Bryan Ehrlich}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We show that a finite-dimensional system admitting a faithful
self-model---an isomorphic copy through which the system can probe and
update its own state---is necessarily governed by complex quantum
mechanics.  Starting from a single operational premise (faithful
self-modeling) and four standing structural assumptions, we construct
a sequential product on
the system's effect algebra using compressions and Peirce
decompositions.  This product satisfies the seven axioms of van de
Wetering, forcing the state space to be a Euclidean Jordan algebra.
The faithful self-modeling constraint further implies local tomography
for the body--model composite, which excludes all non-complex Jordan
types and promotes the algebra to $\Mnsa{n}$ equipped with the
L\"uders sequential product and conjugate-transpose involution.  The
result distinguishes this work from other reconstruction programs,
which typically require 3--6 independent axioms and often assume local
tomography or the complex field.  Our derivation requires a single
operational premise and four standing structural assumptions
(finite-dimensionality, faithfulness, minimal composite, simple EJA).
We argue that these assumptions are not independent design choices but
structural consequences of self-modeling in a finite-capacity system.
\end{abstract}


%% =====================================================================
%%  SECTION 1: INTRODUCTION
%% =====================================================================
\section{Introduction}
\label{sec:intro}

A finite-dimensional physical system that faithfully models itself is
necessarily governed by complex quantum mechanics.  More precisely, if
the state space of a system admits a faithful self-model---an
isomorphic copy through which the system can operationally probe and
update its own state---then the state space is $\Mnsa{n}$ for some
$n$, the algebra of self-adjoint $n \times n$ complex matrices,
equipped with the L\"uders sequential product
$\seqp{a}{b} = \sqrt{a}\, b\, \sqrt{a}$ and the conjugate-transpose
involution $X^{*} = X^{\dagger}$.  No assumption about the complex
field, Hilbert space structure, or $C^{*}$-algebraic operations enters
the premises; all of these emerge as consequences.

This result contributes to the quantum reconstruction program, which
seeks to derive the mathematical framework of quantum theory from
physically transparent principles.  Beginning with Hardy's five-axiom
reconstruction~\cite{Hardy2001}, the program has produced a succession
of results that characterize quantum theory within the landscape of
generalized probabilistic theories (GPTs): Chiribella, D'Ariano, and
Perinotti~\cite{CDP2011} used six axioms including Purification;
Daki\'c and Brukner~\cite{DakicBrukner2011} used four axioms
including local tomography; Masanes and
M\"uller~\cite{MasanesMueller2011,MasanesMueller2013} distilled the
framework to three postulates assuming tomographic locality; and
Barnum, M\"uller, and
Ududec~\cite{BarnumMuellerUdudec2014} used four postulates including
energy observability.  A comprehensive review of the landscape is
given by M\"uller~\cite{Mueller2021}.  A common feature of these
programs is the use of multiple independent operational axioms, and in
most cases the complex number field---or a property that selects it,
such as local tomography---is assumed rather than derived.

Our main result can be stated informally as follows.

\begin{theorem}[Main result, informal]\label{thm:main-informal}
  If a finite-dimensional system's state space admits a faithful
  self-model---an order-isomorphic copy through which the system can
  test effects, update its model, and test again---then the state
  space is $\Mnsa{n}$ for some $n \geq 2$, equipped with the
  L\"uders sequential product and conjugate-transpose involution.
\end{theorem}

The derivation proceeds through a chain of implications:
self-modeling $\to$ sequential product construction $\to$ axiom
verification $\to$ Euclidean Jordan algebra classification $\to$
local tomography $\to$ type exclusion $\to$ $C^{*}$-algebra promotion
$\to$ involution exhibition.
\Cref{fig:chain} illustrates this chain, distinguishing the novel
contributions of this paper from the published theorems invoked.

\Cref{tab:comparison} compares the premise count and complex-selection
mechanism of this work against the major reconstruction programs.

\begin{table}[ht]
\centering
\caption{Comparison of quantum reconstruction programs.  ``Premises''
counts independent operational postulates.  Our premise
requires four standing structural assumptions (see
Assumptions~\ref{ass:finite}--\ref{ass:simple}); maximal coherence
of the Peirce feedback is derived from \ref{ax:S5}, not assumed.}
\label{tab:comparison}
\smallskip
\begin{tabular}{@{}llcl@{}}
\toprule
Program & Year & Premises & Complex selection \\
\midrule
Hardy~\cite{Hardy2001} & 2001 & 5 & Simplicity axiom \\
Chiribella--D'Ariano--Perinotti~\cite{CDP2011} & 2011 & 6 & Purification \\
Daki\'c--Brukner~\cite{DakicBrukner2011} & 2011 & 4 & Local tomography (assumed) \\
Masanes--M\"uller~\cite{MasanesMueller2013} & 2013 & 3 & Tomographic locality (assumed) \\
Barnum--M\"uller--Ududec~\cite{BarnumMuellerUdudec2014} & 2014 & 4 & Energy observability \\
Selby--Scandolo--Coecke~\cite{SelbyScandaloCoecke2021} & 2021 & ${\sim}6$ & Symmetric purification \\
\textbf{This work} & \textbf{2026} & \textbf{1} & \textbf{Derived (LT from faithful tracking)} \\
\bottomrule
\end{tabular}
\end{table}

% Derivation chain figure
\input{figures/derivation-chain}

The paper is organized as follows.
\Cref{sec:prelim} introduces order unit spaces, compressions, Peirce
decompositions, sequential products, and the self-modeling definition.
\Cref{sec:sp} presents the main novel construction: the self-modeling
sequential product, the naive extension and its failure, the corrected
product via Peirce feedback, and the faithful self-modeling selection
principle.
\Cref{sec:axioms} verifies that the corrected product satisfies
axioms S1--S7 of van de Wetering~\cite{vandeWetering2019b}.
\Cref{sec:composite} constructs the body--model composite, proves
local tomography, and addresses the entangled sector.
\Cref{sec:exclusion} excludes all non-complex Euclidean Jordan algebra
types and promotes the algebra to a $C^{*}$-algebra via the theorems
of van de Wetering, Barnum--Wilce, and Hanche-Olsen.
\Cref{sec:discussion} discusses the scope and limitations of the
result.

We close this introduction with an honest statement of scope.  Our
result holds for finite-dimensional systems
(\Cref{ass:finite}), assumes a faithful self-model exists
(\Cref{ass:faithful}), uses the minimal composite construction
(\Cref{ass:minimal}), and restricts to simple Euclidean Jordan
algebras (\Cref{ass:simple}).  The claim is therefore: \emph{one
operational premise plus four standing structural assumptions derive
complex quantum mechanics.}  We argue in \Cref{sec:discussion} that
these structural assumptions are not independent design choices but
consequences of what it means to be a faithful self-modeler in a
finite-capacity system.


%% =====================================================================
%%  SECTION 2: PRELIMINARIES
%% =====================================================================
\section{Preliminaries}
\label{sec:prelim}


\subsection{Order Unit Spaces}
\label{sec:ous}

We work within the framework of order unit spaces, following
Alfsen--Shultz~\cite{AlfsenShultz2003}.

\begin{definition}[Order unit space]\label{def:ous}
  An \emph{order unit space} $(V, V^{+}, \id)$ is a real ordered
  vector space with a proper cone $V^{+}$ (closed, convex, pointed)
  generating $V$, equipped with a distinguished Archimedean order unit
  $\id$: for every $a \in V$ there exists $r \geq 0$ with
  $-r\id \leq a \leq r\id$.  The \emph{order unit norm} is
  $\|a\| = \inf\{r \geq 0 : -r\id \leq a \leq r\id\}$.
\end{definition}

The \emph{effect space} of $V$ is
$\eff{V} = \{a \in V : 0 \leq a \leq \id\}$.  An effect
$p \in \eff{V}$ is \emph{sharp} (a \emph{projective unit}) if $p$ and
$\id - p$ are both extremal in $\eff{V}$, or equivalently if $p$
admits no non-trivial refinement.  We write $\Proj(V)$ for the set of
projective units.

Every effect in a spectral order unit space admits a \emph{spectral
decomposition} $a = \sum_{i} \lambda_{i}\, p_{i}$, where the $p_{i}$
are mutually orthogonal projective units with
$\sum_{i} p_{i} = \id$ and the $\lambda_{i}$ are distinct elements of
$[0,1]$.  This decomposition exists as a property of the OUS lattice
structure~\cite{AlfsenShultz2003}, prior to and independent of any
sequential product.

\begin{assumption}[Finite-dimensional spectral OUS]\label{ass:finite}
  The system's state space is described by a finite-dimensional
  spectral order unit space $(V, V^{+}, \id)$.
\end{assumption}


\subsection{Compressions and Peirce Decomposition}
\label{sec:compressions}

For each projective unit $p \in \Proj(V)$, the Alfsen--Shultz theory
provides a \emph{compression} $\comp{p} : V \to V$, a positive linear
projection satisfying $\comp{p}(\id) = p$.  The compression $\comp{p}$
projects onto the face generated by $p$, and it plays the role of the
L\"uders operation $b \mapsto p\, b\, p$ in the general OUS setting.

\begin{definition}[Peirce decomposition]\label{def:peirce}
  For a projective unit $p \in \Proj(V)$, the \emph{Peirce
  decomposition} of $V$ relative to $p$ is the direct-sum
  decomposition
  \begin{equation}\label{eq:peirce}
    V = V_{2}(p) \oplus V_{1}(p) \oplus V_{0}(p),
  \end{equation}
  where $V_{2}(p) = \mathrm{range}(\comp{p})$ is the face of $p$,
  $V_{0}(p) = \mathrm{range}(\comp{p^{\perp}})$ is the face of
  $p^{\perp} = \id - p$, and
  $V_{1}(p) = \ker(\comp{p}) \cap \ker(\comp{p^{\perp}})$ is the
  \emph{Peirce $1$-space}.
\end{definition}

A key identity is:
\begin{equation}\label{eq:pinching}
  \comp{p} + \comp{p^{\perp}} = \pinch_{p}
  \quad (\text{the pinching map, \emph{not} the identity}).
\end{equation}
The pinching map annihilates $V_{1}(p)$.  In the commutative
(classical) case $V_{1}(p) = \{0\}$, so the pinching map is the
identity; in the non-commutative case $V_{1}(p) \neq \{0\}$ and the
identity fails.  This distinction drives the entire construction in
\Cref{sec:sp}.

More generally, for a resolution of unity
$\{p_{1}, \ldots, p_{n}\}$ (mutually orthogonal projective units with
$\sum_{i} p_{i} = \id$), the \emph{Peirce $(i,j)$ projection}
$\peirce{ij}$ extracts the component of $b$ in the $(i,j)$ Peirce
$1$-space $V_{1}(p_{i}, p_{j})$:
\begin{equation}\label{eq:peirce-proj}
  \peirce{ij}(b) \in V_{1}(p_{i}, p_{j}),
  \qquad
  b = \sum_{i} \comp{p_{i}}(b) + \sum_{i < j} \peirce{ij}(b).
\end{equation}


\subsection{Sequential Products}
\label{sec:sp-axioms}

Following Gudder and Greechie~\cite{GudderGreechie2002} and van de
Wetering~\cite{vandeWetering2019b}, we define a sequential product on
an order unit space.

\begin{definition}[Sequential product space,
  {\cite[Def.~2]{vandeWetering2019b}}]
\label{def:sp}
  Let $(V, \leq, \id)$ be an order unit space equipped with a binary
  operation $(\cdot) : \eff{V} \times \eff{V} \to \eff{V}$.  Write
  $a \compatible b$ (``$a$ and $b$ are compatible'') when
  $\seqp{a}{b} = \seqp{b}{a}$.  Then $(V, \cdot)$ is a \emph{sequential
  product space} if the following axioms hold:
  \begin{enumerate}[label=\textup{(S\arabic*)},leftmargin=3em]
    \item\label{ax:S1} \textbf{Additivity:}
      $\seqp{a}{(b+c)} = \seqp{a}{b} + \seqp{a}{c}$ whenever
      $b + c \leq \id$.
    \item\label{ax:S2} \textbf{Continuity:}
      The map $a \mapsto \seqp{a}{b}$ is continuous in the order unit
      norm.
    \item\label{ax:S3} \textbf{Unitality:}
      $\seqp{\id}{a} = a$ for all $a \in \eff{V}$.
    \item\label{ax:S4} \textbf{Symmetry of orthogonality:}
      If $\seqp{a}{b} = 0$ then $\seqp{b}{a} = 0$.
    \item\label{ax:S5} \textbf{Associativity of compatible effects:}
      If $a \compatible b$ then
      $\seqp{a}{(\seqp{b}{c})} = \seqp{(\seqp{a}{b})}{c}$.
    \item\label{ax:S6} \textbf{Additivity of compatibility:}
      If $a \compatible b$ then $a \compatible (\id - b)$; if also
      $a \compatible c$ then $a \compatible (b + c)$ whenever
      $b + c \leq \id$.
    \item\label{ax:S7} \textbf{Multiplicativity of compatibility:}
      If $a \compatible b$ and $a \compatible c$ then
      $a \compatible \seqp{b}{c}$.
  \end{enumerate}
\end{definition}

Two central results of van de Wetering are used as black boxes
throughout this paper.

\begin{theorem}[{\cite[Thm.~1]{vandeWetering2019b}}]\label{thm:vdw1}
  A finite-dimensional sequential product space is order-isomorphic
  to a Euclidean Jordan algebra.
\end{theorem}

\begin{theorem}[{\cite[Thm.~3]{vandeWetering2019b}}]\label{thm:vdw3}
  If a sequential product space has a locally tomographic composite
  that also carries a sequential product, then the sequential product
  space embeds into a $C^{*}$-algebra.
\end{theorem}

The Jordan--von Neumann--Wigner
classification~\cite{JordanVonNeumannWigner1934} of simple
finite-dimensional Euclidean Jordan algebras yields five types:
$\Mnsa{n}$ over $\mathbb{R}$, $\mathbb{C}$, or $\mathbb{H}$; the
spin factors $\spinfactor{n}$; and the exceptional Albert algebra
$M_{3}(\mathbb{O})\sa$.  Our task is to show that self-modeling
selects the complex type uniquely.


\subsection{Self-Modeling}
\label{sec:self-modeling}

We now introduce the central operational premise.

Consider a physical system $B$ (the ``body'') described by a
finite-dimensional spectral OUS $(V_{B}, V_{B}^{+}, \id_{B})$.  A
\emph{self-model} of $B$ is a second system $M$ (the ``model''),
described by an OUS $(V_{M}, V_{M}^{+}, \id_{M})$, together with a
positive unital map $\varphi: V_{B} \to V_{M}$ (the \emph{tracking
map}) that encodes the operational cycle:
\begin{enumerate}[label=(\roman*)]
  \item \emph{Prepare:} the body $B$ is in some state;
  \item \emph{Update:} testing a sharp effect $p$ on $B$ triggers
    $M$ to undergo the compression $\comp{\varphi(p)}$, updating
    the model;
  \item \emph{Measure:} a second effect $b$ is tested on $B$.
\end{enumerate}
The composite outcome of this cycle---``first $p$, then $b$''---defines
a binary operation on the effect space of $B$.

\begin{assumption}[Faithful self-model]\label{ass:faithful}
  The tracking map $\varphi: V_{B} \to V_{M}$ is an order
  isomorphism.  That is, $M$ is a faithful copy of $B$: it has the
  same dimension, the same order structure, and the same effect space,
  up to relabeling.
\end{assumption}

This is the sole operational premise.  Faithful self-modeling means
the system possesses a complete internal representation of itself, one
that preserves all operationally accessible structure.  Approximate or
partial self-models (where $\varphi$ is merely positive or injective
but not surjective) might yield weaker algebraic structures; we do not
pursue this here.


%% =====================================================================
%%  SECTION 3: THE SELF-MODELING SEQUENTIAL PRODUCT
%% =====================================================================
\section{The Self-Modeling Sequential Product}
\label{sec:sp}


\subsection{From Self-Modeling to Sequential Product}
\label{sec:sp-sharp}

The self-modeling cycle of \Cref{sec:self-modeling} produces a natural
binary operation on the effects of $B$.  For sharp effects
$p, q \in \Proj(V_{B})$, the composite outcome ``first test $p$,
update the model, then test $q$'' is encoded by the compression
product:
\begin{equation}\label{eq:sharp-product}
  \seqp{p}{q} = \comp{p}(q),
\end{equation}
where $\comp{p}$ is the Alfsen--Shultz compression for the face
generated by $p$ in $V_{B}$.

Several properties of this product follow immediately from the
compression axioms.  It maps effects to effects (positivity of
$\comp{p}$).  It satisfies $\seqp{\id}{q} = q$ (since $\comp{\id}$ is
the identity).  On a simplex (classical state space), it reduces to
the pointwise product
$\seqp{p}{q} = (p_{1}q_{1}, \ldots, p_{n}q_{n})$, matching the unique
sequential product on simplices~\cite{GudderGreechie2002}.

\begin{remark}
  The product~\eqref{eq:sharp-product} is defined entirely on
  $V_{B}$ using $B$'s own compression structure.  The tracking map
  $\varphi$ provides the physical justification---testing $p$ on $B$
  causes the model $M$ to undergo $\comp{\varphi(p)}$, which feeds
  back to the context for testing $q$---but the algebraic product is
  intrinsic to $B$.
\end{remark}


\subsection{The Naive Extension and Its Failure}
\label{sec:sp-naive}

To extend the product from sharp to general effects, one attempts a
spectral extension.  Given $a = \sum_{i} \lambda_{i}\, p_{i}$
(spectral decomposition in the OUS), define:
\begin{equation}\label{eq:naive}
  \seqp{a}{b} \stackrel{?}{=}
    \sum_{i} \lambda_{i}\, \comp{p_{i}}(b).
\end{equation}
This is the natural extension: the effect $a$ acts on $b$ as a
weighted sum of compressions, one for each spectral projector.

However, this extension \emph{fails} axiom \ref{ax:S3} (unitality)
on non-commutative order unit spaces.  To see why, set $a = \id$ with
spectral decomposition $\id = \sum_{i} p_{i}$ (all eigenvalues equal
to~$1$).  Then~\eqref{eq:naive} gives
\begin{equation}\label{eq:naive-failure}
  \seqp{\id}{b}
  = \sum_{i} \comp{p_{i}}(b)
  = \pinch_{p_{1}, \ldots, p_{n}}(b)
  \neq b
  \quad\text{in general.}
\end{equation}
The sum of compressions is the \emph{pinching map}, not the identity.
By the Peirce decomposition~\eqref{eq:peirce-proj}, the pinching map
annihilates all Peirce $1$-space components: $\peirce{ij}(b)$ is
discarded for every $i < j$.  In the commutative case
$V_{1} = \{0\}$, the pinching map \emph{is} the identity and the
extension succeeds---this is why the classical limit works.  In the
non-commutative case, the Peirce $1$-space is nontrivial, and the
naive extension destroys off-diagonal information.

This failure is not a bug; it reveals a structural gap.  The
compression sum captures the Peirce $2$-space (diagonal) contributions
correctly but provides no prescription for the Peirce $1$-space
(off-diagonal) terms.  Closing this gap requires new input.


\subsection{The Corrected Product via Peirce Feedback}
\label{sec:sp-corrected}

The missing Peirce $1$-space contribution comes from the self-modeling
feedback loop.  We now show that the form of the corrected product is
\emph{forced} by the axioms, not chosen.  Any binary operation on
effects satisfying \ref{ax:S1} (additivity in the second argument),
\ref{ax:S2} (continuity in the first), and the sharp constraint
$\seqp{p}{b} = \comp{p}(b)$ must respect the Peirce decomposition:
the map $\seqp{a}{\cdot}$ is a linear endomorphism (by \ref{ax:S1}
and finite-dimensionality) whose range on each Peirce subspace is
constrained to that subspace (by the compression algebra structure).
The diagonal coefficients are pinned to the eigenvalues by
eigenvalue coalescence and \ref{ax:S2}.  Thus the general product
for $a = \sum_{i} \lambda_{i}\, p_{i}$ must take the form:
\begin{equation}\label{eq:general-product}
  \seqp{a}{b}
  = \sum_{i} \lambda_{i}\, \comp{p_{i}}(b)
  + \sum_{i < j} f(\lambda_{i}, \lambda_{j})\, \peirce{ij}(b),
\end{equation}
where the first sum captures diagonal (Peirce $2$-space) contributions
via compressions, and the second sum restores the off-diagonal (Peirce
$1$-space) contributions through a \emph{mixing function}
$f: [0,1] \times [0,1] \to \mathbb{R}$ to be determined.

\emph{Why this form is forced.}
By \ref{ax:S1}, $\seqp{a}{\cdot}$ is a linear endomorphism of~$V$.
The Peirce decomposition with respect to the spectral projectors of~$a$
decomposes $V$ into subspaces $V_2(p_i)$ and $V_1(p_i, p_j)$.  Since
compressions are projections onto these subspaces
(Alfsen--Shultz~\cite{AlfsenShultz2003}), linearity forces
$\seqp{a}{\cdot}$ to act by a scalar on each Peirce subspace.  The
sharp constraint $\seqp{p}{b} = \comp{p}(b)$ pins the Peirce $2$-space
scalars to eigenvalues by coalescence (letting $\lambda_i \to \lambda_j$
and applying \ref{ax:S2}).  Thus~\eqref{eq:general-product} is the
unique form compatible with \ref{ax:S1}, \ref{ax:S2}, and the
sharp constraint; only the Peirce $1$-space
scalars~$f(\lambda_i, \lambda_j)$ remain free.  The detailed
computation appears in \Cref{app:proofs}.

The mixing function is constrained by the requirement that the product
map effects to effects: $0 \leq \seqp{a}{b} \leq \id$ for all effects
$a, b$.  This yields a \emph{positivity bound}.

\begin{proposition}[Positivity bound]\label{prop:pos-bound}
  For the product~\eqref{eq:general-product} to map
  $\eff{V} \times \eff{V}$ into $\eff{V}$, the mixing function must
  satisfy
  \begin{equation}\label{eq:pos-bound}
    |f(\lambda_{i}, \lambda_{j})| \leq
    \sqrt{\lambda_{i}\, \lambda_{j}}
  \end{equation}
  for all $\lambda_{i}, \lambda_{j} \in [0,1]$.
\end{proposition}

\begin{proof}
  Every pair of orthogonal atoms $p_{i}, p_{j}$ in a spectral OUS
  generates a two-level face isomorphic to a spin factor, on which
  the Schur complement criterion gives
  $|f(\lambda_{i}, \lambda_{j})| \leq \sqrt{\lambda_{i}\,
  \lambda_{j}}$.  Face effects embed into global effects, so the
  bound lifts to the full space.  The detailed computation appears
  in \Cref{app:proofs}.
\end{proof}

Any mixing function satisfying~\eqref{eq:pos-bound} with
$f(1, \lambda) = \lambda$ and $f(0, \lambda) = 0$ yields a product
that maps effects to effects and satisfies \ref{ax:S3}.  The question
is: \emph{which} $f$ does self-modeling select?


\subsection{Faithful Self-Modeling Selects the Maximum}
\label{sec:sp-faithful}

The faithful self-modeling constraint of \Cref{ass:faithful} selects
the mixing function that saturates the positivity bound.

\begin{proposition}[Maximal coherence]\label{prop:coherence}
  The mixing function saturates the positivity bound:
  \begin{equation}\label{eq:faithful-f}
    f(\lambda_{i}, \lambda_{j}) =
    \sqrt{\lambda_{i}\, \lambda_{j}}.
  \end{equation}
\end{proposition}

\begin{proof}
  Let $f$ be any continuous mixing function satisfying the
  positivity bound~\eqref{eq:pos-bound}.  For compatible
  $a = \alpha_{1} p_{1} + \alpha_{2} p_{2}$ and
  $b = \beta_{1} p_{1} + \beta_{2} p_{2}$ sharing an eigenbasis, \ref{ax:S5}
  requires $\seqp{a}{(\seqp{b}{d})} = \seqp{(\seqp{a}{b})}{d}$ for
  all effects $d$.  Comparing the Peirce $1$-space coefficients of both
  sides gives the multiplicative functional equation
  \begin{equation}\label{eq:func-eq}
    f(\alpha_{1}, \alpha_{2})\, f(\beta_{1}, \beta_{2})
    = f(\alpha_{1}\beta_{1},\; \alpha_{2}\beta_{2}).
  \end{equation}
  With the boundary conditions $f(1, \lambda) = \lambda$
  and $f(0, \lambda) = 0$ (from \ref{ax:S3} and positivity),
  factorization gives $f(x, y) = g(x)\, h(y)$ where $g$ and $h$ are
  continuous multiplicative functions on $[0,1]$.  By the
  multiplicative Cauchy theorem on $\mathbb{R}_{> 0}$, each is a
  power: $g(x) = x^{s}$, $h(y) = y^{t}$.  The boundary condition
  $f(1, \lambda) = \lambda$ forces $t = 1 - s$; symmetry of the
  positivity bound under $(i, j) \leftrightarrow (j, i)$ gives
  $s = t$, hence $s = t = 1/2$.  Therefore
  $f(\lambda_{i}, \lambda_{j}) = \sqrt{\lambda_{i}\, \lambda_{j}}$.
\end{proof}

\noindent
The key step is the multiplicative functional
equation~\eqref{eq:func-eq}: the mixing function must be
\emph{multiplicatively homomorphic} in the eigenvalue pairs.
Combined with continuity (\ref{ax:S2}) and the boundary conditions,
this forces the unique solution
$f = \sqrt{\lambda_{i}\, \lambda_{j}}$.  Note that maximal coherence
is not an additional assumption: it is forced by the sequential
product axioms.  The role of self-modeling in this result is
\emph{constructive}: the tracking isomorphism~$\varphi$ provides
the sequential product on which the axioms act
(\Cref{sec:sp-def}).  Without~$\varphi$, there is no product to
constrain; once the product exists, the axioms determine~$f$
uniquely.

The corrected product with
$f = \sqrt{\lambda_{i}\, \lambda_{j}}$ is:
\begin{equation}\label{eq:corrected-product}
  \boxed{
  \seqp{a}{b}
  = \sum_{i} \lambda_{i}\, \comp{p_{i}}(b)
  + \sum_{i < j} \sqrt{\lambda_{i}\, \lambda_{j}}\,
    \peirce{ij}(b).
  }
\end{equation}

\begin{remark}
  On $\Mnsa{2}$, the corrected
  product~\eqref{eq:corrected-product} coincides with the L\"uders
  product $\seqp{a}{b} = \sqrt{a}\, b\, \sqrt{a}$.  This equivalence
  is a \emph{consequence} of the construction, not a premise.  The
  L\"uders formula was never assumed; it emerges from the positivity
  bound and faithful self-modeling.
\end{remark}

\begin{remark}
  The role of the tracking map $\varphi$ is now clear.  In the sharp
  product~\eqref{eq:sharp-product}, $\varphi$ entered only as
  physical interpretation.  In the corrected product, $\varphi$ is
  algebraically essential: setting $f = 0$ (trivial model, no
  feedback) recovers the failed naive extension, while setting
  $f = \sqrt{\lambda_{i}\, \lambda_{j}}$ (faithful model, maximal
  feedback) gives the corrected product.  Self-modeling quality
  controls coherence.
\end{remark}

The remaining structural assumptions are:

\begin{assumption}[Minimal composite]\label{ass:minimal}
  The body--model composite $V_{BM}$ is the minimal order unit space
  carrying product states, product effects, non-signaling constraints,
  and a product-form sequential product.
\end{assumption}

\begin{assumption}[Simple EJA]\label{ass:simple}
  The Euclidean Jordan algebra produced by \Cref{thm:vdw1} is simple
  (it has no nontrivial Jordan ideal).
\end{assumption}


\subsection{Circularity Check}
\label{sec:circularity}

A natural concern is whether the scalar function
$\sqrt{\lambda_{i}\, \lambda_{j}}$
in~\eqref{eq:corrected-product} imports $C^{*}$-algebraic structure
through the back door.  It does not.  The square root in the mixing
function operates on \emph{real eigenvalues}
$\lambda_{i}, \lambda_{j} \in [0,1]$: it is arithmetic, specifically
$\sqrt{x} : [0,1] \to [0,1]$, the unique non-negative real number
whose square is $x$.  No operator square root, no trace, no inner
product, no density matrix, and no Hilbert space structure is used in
the construction.  The only mathematical objects employed are:
\begin{itemize}
  \item the order unit space $(V, V^{+}, \id)$;
  \item the effect space $\eff{V}$ and its projective units;
  \item Alfsen--Shultz compressions $\comp{p}$ for faces of
    projective units;
  \item the Peirce decomposition~\eqref{eq:peirce-proj} derived from
    compressions;
  \item spectral decomposition of effects in the OUS;
  \item the scalar function $\sqrt{\cdot}$ on
    $[0,1] \subset \mathbb{R}$.
\end{itemize}
Complex numbers, Hilbert spaces, and $C^{*}$-algebraic operations do
not appear until the conclusion (\Cref{sec:exclusion}), where they
emerge as consequences of the classification theorems.


%% =====================================================================
%%  SECTIONS 4--6: BODY (from Plan 02)
%% =====================================================================

\input{sections/axiom-verification}

\input{sections/composite-lt}

\input{sections/type-exclusion}


%% =====================================================================
%%  SECTION 7: DISCUSSION (from Plan 03)
%% =====================================================================

\input{sections/discussion}


%% =====================================================================
%%  ACKNOWLEDGMENTS
%% =====================================================================

\section*{Acknowledgments}

[To be added.]


%% =====================================================================
%%  APPENDICES
%% =====================================================================
\appendix

\input{sections/appendix-proofs}

\input{sections/appendix-numerical}


%% =====================================================================
%%  BIBLIOGRAPHY
%% =====================================================================
\bibliographystyle{alpha}
\bibliography{refs}

\end{document}
