% main.tex — Paper 0: Radical Relativity
\documentclass[12pt]{article}

% --- Packages ---
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage{float}
\usepackage{natbib}

% --- Hyperref ---
\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  citecolor=green!40!black,
  urlcolor=blue!70!black
}

% --- Theorem-style environments (philosophy, not proofs) ---
\newtheorem{postulate}{Postulate}
\newtheorem{definition}{Definition}
\newtheorem{claim}{Claim}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

% --- Notation ---
\newcommand{\hthree}{h_3(\mathbb{O})}
\newcommand{\Ffour}{F_4}
\newcommand{\SMgauge}{(U(1) \times SU(2) \times SU(3))/\mathbb{Z}_6}
\newcommand{\Mnsa}[1]{M_{#1}(\mathbb{C})^{sa}}
\newcommand{\rhoJ}{\rho_J}
\newcommand{\SU}[1]{\mathrm{SU}(#1)}
\newcommand{\UU}[1]{\mathrm{U}(#1)}
\newcommand{\Spin}[1]{\mathrm{Spin}(#1)}

% --- Title ---
\title{Radical Relativity: A Manifesto for Physics\\from the Absence of Privileged Instantiation}
\author{Bryan Ehrlich}
\date{}

\begin{document}
\maketitle

% ====================================================================
% ABSTRACT
% ====================================================================

\begin{abstract}
The Copernican program removed privileged positions, velocities, frames,
and coordinate systems from physics.  One distinction remains: the
distinction between ``mathematical'' and ``physical.''  We argue that
removing it --- refusing to privilege any instantiation of a mathematical
structure over any other --- completes the program and forces observable
physics.  The philosophical premise is \emph{radical relativity}: no
instantiation of a consistent mathematical structure is ontologically
privileged.

From this premise, together with one operational requirement (faithful
self-modeling) and four structural assumptions argued to be consequences
of self-modeling in a finite-capacity system, we summarize a derivation
chain connecting structure-space ontology to complex quantum mechanics,
general relativity, and the Standard Model gauge group.  Quantum
mechanics follows by theorem from the self-modeling premise and its
structural assumptions: the state space must carry $C^*$-algebraic
structure with complex field, Born rule, and unitary dynamics (companion
Paper~5).  General relativity follows conditionally from locality of
self-modeling via the Jacobson thermodynamic route (companion Paper~6).
The Standard Model gauge group emerges from the exceptional Jordan
algebra $\hthree$, the unique non-composable self-modeling structure
(companion Paper~7).  The experiential measure resolves uniquely
within $\hthree$: $\rhoJ = \det(X)\,(\mathrm{Tr}(X^2) -
\tfrac{1}{3})$ is the unique lowest-degree $\Ffour$-invariant
satisfying the self-modeling boundary conditions, producing a
falsifiable cubic-vs-quadratic prediction that distinguishes the
framework from integrated information theory (companion Paper~1,
revised).

This paper is a manifesto and roadmap.
Its purpose is to present the philosophical case, summarize the
derivation chain, register the open gaps honestly, and situate the
program relative to Tegmark, Mueller, Wolfram, and Connes.
\end{abstract}

% ====================================================================
% AUTHORSHIP NOTE
% ====================================================================

\section*{A note on authorship}

I am a software engineer, not a mathematician or physicist.  AI tools
assisted with formal exposition; the ideas and editorial decisions are
mine.  Feedback on the philosophy and mathematics is especially welcome.

% ====================================================================
% SECTIONS
% ====================================================================

\input{sections/introduction}
\input{sections/why-anything}
\input{sections/why-self-modelers}
\input{sections/why-qm}
\input{sections/why-this-physics}
\input{sections/objections}
\input{sections/research-program}

% ====================================================================
% BIBLIOGRAPHY
% ====================================================================

\bibliographystyle{unsrtnat}
\bibliography{refs}

\end{document}
