% Appendix B: Numerical Verification
% \input-able LaTeX file for Paper 5

\section{Numerical Verification}\label{app:numerics}

All results in this paper have been verified by an automated symbolic
test suite implemented in SymPy~1.14~\cite{SymPy2017}.  The test suite
uses exact rational arithmetic (no floating-point approximations) and
is available as supplementary material
(\texttt{code/sp\_verification.py}).

%--------------------------------------------------------------------
\subsection{Test Suite Overview}

The verification was conducted in two phases, corresponding to the
two main stages of the derivation.

\paragraph{Phase~1: Sequential product axioms (S1--S7).}
Working on $\Mnsa{2}$ (the qubit system, identified with the spin
factor $\spinfactor{3}$), we verified that the corrected sequential
product~\eqref{eq:corrected-product} satisfies all seven axioms of
van de Wetering~\cite{vandeWetering2019b}.  The test suite includes
positive controls (the L\"uders product $\seqp{a}{b} = \sqrt{a}\,b\,
\sqrt{a}$ passes all axioms) and negative controls (matrix
multiplication fails~\ref{ax:S4}).
A total of \textbf{186~tests} were executed:

\begin{center}
\begin{tabular}{@{}llr@{}}
\toprule
Category & Description & Tests \\
\midrule
S1 (additivity)      & Linearity in second argument       & 20 \\
S2 (continuity)      & Parametric path, 11-point sweep     & 11 \\
S3 (unitality)       & $\seqp{\id}{a} = a$ for 14 effects    & 14 \\
S4 (orth.\ symmetry) & Sharp, general, rotated effects     & 186$^{*}$ \\
S5 (comp.\ assoc.)   & Compatible triples                  & 24 \\
S6 (comp.\ add.)     & Complement + sum tests              & 8 \\
S7 (comp.\ mult.)    & Functional calculus                  & 6 \\
Non-associativity    & Explicit witness + 20 random triples & 22 \\
Classical limit      & Pointwise product on simplices       & 25 \\
L\"uders comparison  & Corrected $=$ L\"uders on $\Mnsa{2}$ & 6 \\
$\varphi$-dependence & S4 under coarse-graining/trivial $\varphi$ & 18 \\
Effect range         & $0 \leq \seqp{a}{b} \leq \id$         & 49 \\
\bottomrule
\end{tabular}
\end{center}

\noindent ${}^{*}$The S4 test count (186) includes sharp orthogonal
projections (10), general effects (12), general-position rotated Bloch
vectors (54), full-rank effects (4), parametric search (52), scaled
projections (30), multiple $\varphi$ choices (13), trivial $\varphi$
(5), and L\"uders positive control (6).  Some tests appear in multiple
categories.

\paragraph{Phase~2: Composite system and type exclusion.}
Working on $\spinfactor{3} \otimes \spinfactor{3}$ (the body--model
composite for qubits), we verified the product-form sequential
product, local tomography, and the type exclusion argument.
A total of \textbf{658+ tests} were executed:

\begin{center}
\begin{tabular}{@{}llr@{}}
\toprule
Category & Description & Tests \\
\midrule
Composite S1--S7 & Axiom inheritance on $\spinfactor{3}
                    \otimes \spinfactor{3}$ & 22 \\
Composite basics  & Product-form SP, unitality, orthogonality & 11 \\
Dimension counting & $4 \times 4 = 16 = \dim(\Mnsa{4})$ & 3 \\
Negative checks    & $\mathbb{R}$: $9 \neq 10$;
                     $\mathbb{H}$: $36 \neq 28$ & 2 \\
Classical limit    & $25 \times 25 = 625$ product pairs on simplex & 625 \\
Involution P1--P4  & $(X^{*})^{*} = X$, $(XY)^{*} = Y^{*}X^{*}$,
                     etc.\ on $\Mnsa{2}$ & 4 \\
Dim.\ formulas     & $\mathbb{R}$, $\mathbb{C}$, $\mathbb{H}$
                     at $n = 2, 3, 4$ & 9 \\
\bottomrule
\end{tabular}
\end{center}

%--------------------------------------------------------------------
\subsection{Summary}

In total, \textbf{844+ automated symbolic tests} were executed across
both phases.  All tests pass.  No regressions were observed across the
sequential execution of the five-plan Phase~4 and two-plan Phase~5
pipelines.

The test suite is designed as a regression harness: any modification
to the sequential product formula, the composite construction, or the
type exclusion arguments can be immediately checked against the full
battery of tests.  The use of exact SymPy arithmetic (rational numbers
and algebraic expressions involving $\sqrt{3}$, $\sqrt{2}$, etc.)
ensures that all verifications are exact, not approximate.
